---
title: "R Markdown Data Analysis of Cyclistic"
output: 
  html_document: 
    highlight: espresso
    theme: cerulean
---

# **LOAD PACKAGES**

#### **TIDYVERSE**

Load the tidyverse package as it includes the packages dplyr and readr which contains the function read_csv()

```{r load tidyverse, message=TRUE, warning=FALSE}

library(tidyverse)

```


#### **LUBRIDATE**

Load the lubridate package to work with dates.

```{r load lubridate, message=FALSE, warning=FALSE}

library(lubridate)

```


#### **SCALES**

Load the scales package to work with plots.

```{r load scales, message=FALSE, warning=FALSE}

library(scales)
```

#### **KNITR**

Load the knitr package to work with 
R Markdown documents and to use the function kable()

```{r load knitr, message=FALSE, warning=FALSE}

library(knitr)

```


# **IMPORT DATASET**

The SQL Server table named "full_year_2021_V3" contained the Cyclistic dataset after data wrangling & data cleaning.  This dataset was then exported as a CSV file to the desktop computer.  This file is now being imported into R as a data frame titled "cyclistic_data".

```{r import csv}
cyclistic_data <- read_csv("sqlexport.csv") 

```

# **PREVIEW DATA FRAME**

Preview the imported Cyclistic dataset such as the
column names and data types by using the function glimpse().

```{r preview data frame, echo=TRUE}

glimpse(cyclistic_data)

```

Function str() displays the structure of the cyclistic dataset including the class and data type of each column.

```{r preview structure}
str(cyclistic_data)

```


# **START STATIONS FREQUENTLY USED**

This Stacked Bar Chart displays the Top 10 start stations most frequently used by both casual & member riders.  Some start stations are used solely by member riders or solely by casual riders.  The X axis of this chart shows the start staton names while the Y axis displays the number of rides for each start station.

```{r popular start stations}
riders_start_stations <-
  cyclistic_data[! is.na(cyclistic_data$start_station_id),] %>% 
  count(member_casual,start_station_id, start_station_name) %>%
  arrange(-n) %>% 
  filter(n >= 15725) 

```



```{r }

knitr::kable(riders_start_stations, "pipe")


```




```{r plot start stations}

ggplot(data=riders_start_stations,
       mapping= aes(x=reorder(start_station_name, -n), y=n, 
                    fill=member_casual))+
  geom_bar(stat="identity") +
  theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1))+
  labs(title= "Popular Start Stations", x="Start Stations",
       y="Number of Rides")+
  scale_y_continuous(breaks=seq(0,90000,10000))+
  scale_fill_manual(values=c("navyblue", "cornflowerblue"))


```



# **BIKE ROUTES MOST FREQUENTLY USED BY RIDERS**

There are 246,265 bike routes.  A route is defined as the pairing of a start station and end station for each bike ride.  This bar chart shows the most popular bike routes with the highest number of rides.

```{r}
cyclistic_data2 <- 
  cyclistic_data %>% 
  drop_na() %>% 
  select(start_station_id, end_station_id, start_station_name, 
         end_station_name, member_casual) %>% 
  filter(start_station_id != end_station_id) %>% 
  unite ("id_start_end", 
         start_station_id, end_station_id, sep = "-") %>% 
  unite ("name_start_end", 
         start_station_name, end_station_name, sep = "-") %>%
  count(id_start_end, name_start_end, member_casual) %>% 
  arrange(-n)

```


```{r}

cyclistic_data3 <- cyclistic_data2 %>% 
                   top_n(10)

```


```{r}
knitr::kable(cyclistic_data3, "pipe")
```


```{r}

ggplot(data=cyclistic_data3,
       mapping= aes(x=reorder(id_start_end, -n), y=n, 
                    fill=id_start_end))+
  geom_bar(stat="identity") +
  theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1))+
  labs(title= "Popular Bike Routes", x="Bike Routes",
       y="Number of Rides")+
  scale_y_continuous(breaks=seq(0,6000,500))+
  annotate("text", x="13008-13022", y=5000, label="Route Lake Shore- Streeter Dr
           Highest Rides", color="red", fontface="bold", size=3)

```



# **SEASON WITH HIGHEST NUMBER OF RIDES**

This Scatter Plot displays number of rides by month for both casual and member riders.  The summer season
in Chicago, June to September had the highest number of rides.  This is highlighted in the chart by creating an annotation of a yellow rectangle.

The data type of column started_datetime is POSIXct and Double which is a DateTime

```{r}
class(cyclistic_data$started_datetime)

typeof(cyclistic_data$started_datetime)
```

Data frame copy is created called cyclistic_data5

```{r}

cyclistic_data5 <- cyclistic_data
```

Month name for column started_datetime is included
in the data frame

```{r}

cyclistic_data5$month_name <- 
  month(ymd_hms(cyclistic_data5$started_datetime),
        label=TRUE, abbr=FALSE)
```


```{r}
season_rides <- cyclistic_data5 %>%
                count(month_name, member_casual) %>% 
                arrange(-n) 

```


```{r}
knitr::kable(season_rides, "pipe")
```


```{r}
ggplot(data=season_rides)+
  geom_point(mapping =aes(x=month_name, y=n, color=member_casual,
                          shape=member_casual))+
  geom_smooth(mapping =aes(x=month_name, y=n))+
  theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1))+
  labs(title= "Seasons - Highest Rides", x="Months",
       y="Number of Rides")+
  scale_y_continuous(labels=comma) +
  annotate("rect", xmin="June", xmax="September", ymin=340000, ymax=450000,
           fill="yellow", color="yellow", alpha=0.3)
```


# **DAYS WITH HIGHEST NUMBER OF RIDES**

This Scatter Plot displays number of rides taken for each day of the week during the year 2021 for
both casual and member riders.

The name of the day for each date in the column started_datetime has been included in the data frame.

The size and shape of the data points for member riders are different from the casual riders.

```{r}
cyclistic_data5$day_name <- 
  wday(ymd_hms(cyclistic_data5$started_datetime),
        label=TRUE, abbr=FALSE)

```

```{r}
day_rides <- cyclistic_data5 %>%
             count(day_name, member_casual) %>% 
             arrange(-n) 
```

```{r}
knitr::kable(day_rides, "pipe")
```

```{r warning=FALSE}
ggplot(data=day_rides)+
  geom_point(mapping =aes(x=day_name, y=n, color=member_casual,
                          shape=member_casual, size=member_casual))+
  theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1))+
  labs(title= "Days - Highest Rides", x="Days",
       y="Number of Rides")+
  scale_y_continuous(labels=comma)
```



# **TIMES WITH HIGHEST NUMBER OF RIDES**

This Scatter Plot displays number of rides by 
the time of day for both casual and member riders, during the year 2021.

Time by hour for column started_datetime is included
in the data frame.

The geom_smooth function is used to show the regression line and patterns in the data.

The facet_wrap function is used to display separate plots for member and casual riders which are subsets of the main dataset.  

```{r}
cyclistic_data5$hours <- 
  hour(ymd_hms(cyclistic_data5$started_datetime))
```

```{r}
time_of_rides <- cyclistic_data5 %>%
                 count(hours, member_casual) %>% 
                 arrange(-n) 
```

```{r}
knitr::kable(time_of_rides, "pipe")
```

```{r}
ggplot(data=time_of_rides)+
  geom_point(mapping =aes(x=hours, y=n, color=member_casual,
                          shape=member_casual))+
  geom_smooth(mapping =aes(x=hours, y=n),
              method="gam", formula=y~s(x))+
  theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1))+
  labs(title= "Time of Day - Highest Rides", x="Time",
       y="Number of Rides")+
  scale_y_continuous(labels=comma)+
  facet_wrap(~member_casual)
```

